Constraints on Cosmic-Ray Particle Populations, Magnetic Fields, and Dark Matter from VHE Gamma-Ray Observations of the Coma Cluster of Galaxies with VERITAS

### 2011-07-05 ms.tex
Fermi section added

%\title{VHE Gamma-Ray Observations of the Coma Cluster of Galaxies with VERITAS}
\title{Constraints on Cosmic-Ray Particle Populations, Magnetic Fields, and Dark Matter from VHE Gamma-Ray Observations of the Coma Cluster of Galaxies with VERITAS}
\author{N. Karlsson\altaffilmark{1,2}, M. Pohl\altaffilmark{3,4}}
\affil{(the VERITAS collaboration)} 
%%\email{nkarlsson@physics.umn.edu}
\and
\author{C. Pfrommer\altaffilmark{5,6},
A. Pinzke\altaffilmark{7}}

@@

\begin{abstract}
Theoretical models of galaxy clusters suggest they may be sources of very high-energy (VHE) gamma-ray emission. Observations of radio halos and relics in clusters indicate populations of relativistic electrons and positrons, which are either injected into the intra-cluster medium and accelerated by turbulent shocks or secondaries from interactions of cosmic-ray hadrons (protons and ions) with the intra-cluster gas. Those interactions would also produce gamma rays through the decay of neutral pions. Reported here are observations of the Coma cluster of galaxies with the VERITAS array of imaging Cherenkov telescopes. The observations were made between March and May 2008. Radio observations at 1.41 GHz taken with the Green Bank Telescope and X-ray observations with the ROSAT All-Sky survey was used to a priori select regions of interest in the field of view. No significant VHE gamma-ray emission from the Coma cluster was detected by VERITAS. Flux upper limits at the 99\% confidence level were obtained for the cluster core, both as a point source and a mildly extended source centered at the cluster core, and for a select few other cluster members. The gamma-ray flux upper limits have been used to constrain the populations of cosmic-ray particles and magnetic fields in the cluster. The cosmic-ray pressure is... The observations imply a lower limit on the intra-cluster magnetic field of X$\mu$G. Finally, a constraint on the dark matter annihilation cross section is derived, $\expval{\sigma v}=$.
\end{abstract}

@@

@@

The data analysis was performed following the standard VERITAS procedures described in \citet{inproc:Cogan_etal:2007} and \citet{inproc:Daniel_etal:2007}. Prior to event reconstruction and selection, all shower images are calibrated and cleaned. Showers are then reconstructed for events with at least two contributing telescope images that pass the following quality selection criteria: more than four participating pixels in the camera, number of photoelectrons in the image is larger than 75, and the distance from the image centroid to the center of the camera is less than $1.43^{\circ}$. These quality selection criteria impose an energy threshold\footnote{Defined as$\ldots$} of about 220 GeV. In addition, events for which only images from the two closest-spaced telescopes\footnote{In the array configuration prior to summer 2009, two telescopes had a separation of only 35 m.} survive quality selection are rejected, as they introduce an irreducible high background rate, due to local muons, degrading the instrument sensitivity \citep{article:MaierKnapp:2007}.

@@

The Coma cluster is a very rich cluster of galaxies with many plausible sites for gamma-ray emission. The core region, the peripheral radio relic, as well as individual powerful cluster member galaxies. VERITAS has a large enough field of view to allow investigation of several of these scenarios. In this work the focus has been on the core region and three cluster members. The latter are treated as point sources in the analysis, whereas the core is treated as both a point source and a mildly extended source. The intrinsic source extension is assumed to be $0.2^{\circ}$ and $0.4^{\circ}$, based on the extension of the thermal soft X-ray emission from the core. There is evidence of a recent merger event between the two central galaxies NGC 4889 and NGC 4874 \citep{article:Tribble:1993}. There is also evidence for an excess of non-thermal X-ray emission from these galaxies as well as the galaxy NGC 4921. Therefore, searches for point-like VHE gamma-ray emission have been conducted at the locations of these galaxies. The regions of interest considered in this work are summarized in Table \ref{table:roi}.

The ring background model \citep{article:Aharonian_etal:2001} is used to estimate the background due to CRs. The total number of events in a given region of interest is then compared to the estimated background from the off-source region scaled by the ratio of the solid angles to produce a final excess. The VHE gamma-ray significance is then calculated using the method by \citet{article:LiMa:1983}.  Significance skymaps over the VERITAS field of view produced with a $0.2^{\circ}$ sampling radius are shown in Figure \ref{fig:skymaps} with overlaid X-ray and radio contours from the ROSAT all-sky survey \citep{article:BrielHenryBohringer:1992}  and GBT 1.4 GHz observations \citep{article:BrownRudnick:2010} respectively.

\section{Fermi-LAT Observations and Analysis}

@@

Table \ref{table:results} lists the upper limits for the selected regions of interest tabulated in Table \ref{table:roi}. These upper-limit calculations depend on the gamma-ray spectrum, which in this work is assumed to be a power law with a spectral index $\alpha=-2.1, -2.3$, and $-2.5$. A power-law spectrum with index $\alpha=-1.5$ is also considered for the dark matter constraints.

@@

\begin{figure*}
\begin{center}
\scalebox{0.45}{\plotone{f1a.eps}}
\scalebox{0.45}{\plotone{f1b.eps}}
\end{center}
\caption{\emph{Left}: Smoothed significance map of the Coma cluster calculated from the observed excess VHE gamma-ray events over a $4.5^{\circ}\times 4.5^{\circ}$ field of view. The excess counts were derived using a ring background model \citep{article:Aharonian_etal:2001}. White contours show the X-ray counts per second in the 0.1 to 2.4 keV energy band from the ROSAT all-sky survey \citep{article:BrielHenryBohringer:1992}. \emph{Right}: Same as above but with overlaid contours from point source subtracted GBT radio observations at 1.4 GHz \citep{article:BrownRudnick:2010}. Shown are also the $0.2^{\circ}$ and $0.4^{\circ}$ radii (dashed cyan) considered for the extended source analyses presented here.}
\label{fig:skymaps}
\end{figure*}

\begin{figure*}
\begin{center}
\scalebox{1.0}{\plotone{f2.eps}}
\end{center}
\caption{$\theta^{2}$ distribution from VERITAS observations of the Coma cluster of galaxies. The points with error bars represent the ON-source data sample and and the filled area the background sample. Each bin represents an annulus around the Coma cluster core position. The data were derived from the ring background model using a $0.2^{\circ}$ sampling radius.}
\label{fig:thetasq}
\end{figure*}

@@

\begin{deluxetable}{lcc}
\tablecolumns{3}
\tablewidth{0pc}
\tablecaption{Regions of interest in the Coma cluster field of view. The cluster core is considered both as a point source and a modestly extended source. Three central galaxies are also considered in point-source searches. The choice is based on evidence of an excess of non-thermal X-ray emission \citep{article:Neumann_etal:2003} at the location of these galaxies.}
\tablehead{
\colhead{Source} &
\colhead{RA} &
\colhead{Dec}}
\startdata
Core & \RA{12}{59}{48.7} & \Dec{+27}{58}{50.0}\\
NGC 4889 & \RA{13}{00}{08.13} & \Dec{+27}{58}{37.03}\\
NGC 4874 & \RA{12}{59}{35.71} & \Dec{+27}{57}{33.37}\\
NGC 4921 & \RA{13}{01}{26.12} & \Dec{+27}{53}{09.59}\\
\enddata
\label{table:roi}
\end{deluxetable}

\begin{landscape}
\begin{deluxetable}{lcccccccccccc}
\tablecolumns{6}
\tablewidth{0pc}
\tablecaption{Flux upper limits for regions of interest in the Coma cluster of galaxies and surroundings.}
\tablehead{
\colhead{Source} & \colhead{RoI\tablenotemark{a} [deg]} & \colhead{$N_{S}$} & \colhead{S\tablenotemark{b} [$\sigma$]} &
\multicolumn{6}{c}{Flux U.L.\tablenotemark{c}} &
\multicolumn{3}{c}{Minimum gamma-ray flux\tablenotemark{d}}\\
\colhead{} & \colhead{} & \colhead{} & \colhead{} & \multicolumn{2}{c}{$\alpha=-2.1$} &
\multicolumn{2}{c}{$\alpha=-2.3$} &
\multicolumn{2}{c}{$\alpha=-2.5$} &
\colhead{$\alpha=-2.1$} &
\colhead{$\alpha=-2.3$} &
\colhead{$\alpha=-2.5$}}
\startdata
Core & 0 & 17 & 0.84 & 2.59 & (0.78\%) & 2.78 & (0.83\%) & 2.97 & (0.89\%) \\
& 0.2 & -41 & -1.0 & 1.96 & (0.59\%)  & 2.09 & (0.63\%) & 2.21 & (0.66\%) \\
& 0.4 & -26 & -0.30 & 4.44 & (1.3\%)  & 4.74 & (1.4\%) & 5.02 & (1.5\%)\\
NGC 4889 & 0 & 3 & 0.14 & - & - & 1.85 & (0.55\%)  & - & - \\
NGC 4874 & 0 & -14 & -0.71 & - & - &  1.51 & (0.45\%)  & - & - \\
NGC 4921 & 0 & -4 & -0.23 & - & - &  2.41 & (0.72\%)  & - & -
\enddata
\tablenotetext{a}{Intrinsic source radius (zero means point source), which is convolved with the analysis PSF.}
\tablenotetext{b}{Statistical significance calculated according to \citet{article:LiMa:1983}.}
\tablenotetext{c}{99\% confidence level upper limit in units of $10^{-8}$ ph. m$^{-2}$ s$^{-1}$ calculated according to \citet{article:Rolke_etal:2005} above an energy threshold of 220 GeV. Values in parentheses are the corresponding fluxes in percent of the steady Crab Nebula flux.}
\tablenotetext{d}{Minimum gamma-ray flux in units of $10^{-13}$ ph. cm$^{-2}$ s$^{-1}$ from the hadronic model described in the text.}
\label{table:results}
\end{deluxetable}
\end{landscape}

>>>>

%\title{VHE Gamma-Ray Observations of the Coma Cluster of Galaxies with VERITAS}
\title{Constraints on Cosmic-Ray Particle Populations, Magnetic Fields, and Dark Matter from VHE Gamma-Ray Observations of the Coma Cluster of Galaxies with VERITAS}
\author{N. Karlsson\altaffilmark{1,2}, M. Pohl\altaffilmark{3,4}}
\affil{(the VERITAS collaboration)} 
%%\email{nkarlsson@physics.umn.edu}
\and
\author{S. Federici\altaffilmark{3,4}, C. Pfrommer\altaffilmark{5,6},
A. Pinzke\altaffilmark{7}}

@@

\begin{abstract}
Theoretical models of galaxy clusters suggest they may be sources of very high-energy (VHE) gamma-ray emission. Observations of radio halos and relics in clusters indicate populations of relativistic electrons and positrons, which are either injected into the intra-cluster medium and accelerated by turbulent shocks or secondaries from interactions of cosmic-ray hadrons (protons and ions) with the intra-cluster gas. Those interactions would also produce gamma rays through the decay of neutral pions. Reported here are observations of the Coma cluster of galaxies with the VERITAS array of imaging Cherenkov telescopes. The observations were made between March and May 2008. Radio observations at 1.41 GHz taken with the Green Bank Telescope and X-ray observations with the ROSAT All-Sky survey was used to a priori select regions of interest in the field of view. No significant VHE gamma-ray emission from the Coma cluster was detected by VERITAS. Several emission regions are considered, including the cluster core as both unresolved (point-like) and spatially extended, and a select few cluster members. The flux upper limits at the 99\% confidence level measured are on the order of $2-5\times 10^{-8}$ ph. m$^{-2}$ s$^{-1}$, which is X times higher than the model-predicted minimum gamma-ray flux.

The gamma-ray flux upper limits have been used to constrain the populations of cosmic-ray particles and magnetic fields in the cluster. The relative cosmic-ray pressure $X_{\mathrm{CR}}$ is. The observations imply a lower limit on the intra-cluster magnetic field of X$\mu$G. Finally, a constraint on the dark matter annihilation cross section is derived, $\expval{\sigma v}=$.
\end{abstract}

@@

A third mechanism for gamma-ray production in a galaxy cluster is self-annihilation of a dark matter particle, e.g. a weakly interacting massive particle (WIMP).

@@

The data analysis was performed following the standard VERITAS procedures described in \citet{inproc:Cogan_etal:2007} and \citet{inproc:Daniel_etal:2007}. Prior to event reconstruction and selection, all shower images are calibrated and cleaned. Showers are then reconstructed for events with at least two contributing telescope images that pass the following quality selection criteria: more than four participating pixels in the camera, number of photoelectrons in the image is larger than 75, and the distance from the image centroid to the center of the camera is less than $1.43^{\circ}$. These quality selection criteria impose an energy threshold\footnote{Defined as the energy corresponding to the maximum of the product function of the observed spectrum with the collection area} of about 220 GeV. In addition, events for which only images from the two closest-spaced telescopes\footnote{In the array configuration prior to summer 2009, two telescopes had a separation of only 35 m.} survive quality selection are rejected, as they introduce an irreducible high background rate, due to local muons, degrading the instrument sensitivity \citep{article:MaierKnapp:2007}.

@@

The Coma cluster is a very rich cluster of galaxies with many plausible sites for gamma-ray emission: the core region, the peripheral radio relic, as well as individual powerful cluster member galaxies. VERITAS has a large enough field of view to allow investigation of several of these scenarios. In this work the focus has been on the core region and three cluster members. The latter are treated as point sources in the analysis, whereas the core is treated as both a point source and a mildly extended source. The intrinsic source extension is assumed to be $0.2^{\circ}$ and $0.4^{\circ}$, based on the extension of the thermal soft X-ray emission from the core. There is evidence of a recent merger event between the two central galaxies NGC 4889 and NGC 4874 \citep{article:Tribble:1993}. There is also evidence for an excess of non-thermal X-ray emission from these galaxies as well as the galaxy NGC 4921 \citep{article:Neumann_etal:2003}. Therefore, searches for point-like VHE gamma-ray emission have been conducted at the locations of these galaxies. The regions of interest considered in this work are summarized in Table \ref{table:roi}.

The ring background model \citep{article:Aharonian_etal:2001} is used to estimate the background due to CRs misinterpreted as gamma rays (the cuts described above reject more than 99\% of all CRs). The total number of events in a given region of interest is then compared to the estimated background from the off-source region scaled by the ratio of the solid angles to produce a final excess. The VHE gamma-ray significance is then calculated using the method by \citet{article:LiMa:1983}.  Significance skymaps over the VERITAS field of view produced with a $0.2^{\circ}$ sampling radius are shown in Figure \ref{fig:skymaps} with overlaid X-ray and radio contours from the ROSAT all-sky survey \citep{article:BrielHenryBohringer:1992}  and GBT 1.4 GHz observations \citep{article:BrownRudnick:2010} respectively.

\section{Fermi-LAT Observations and Analysis}
The Large Area Telescope (LAT) on board the \emph{Fermi} Gamma-ray Space Telescope (\emph{Fermi}) has observed the Coma cluster in all-sky survey mode since its launch in June 2008. The \emph{Fermi}-LAT is sensitive to gamma rays in the 20 MeV to $\sim$300 GeV energy range and is thus very complimentary to the VERITAS obervations. \citet{article:Ackermann_etal:2010} reported on the search for gamma-ray emission from thirty-three galaxy clusters in the data from the first 18 months, including the Coma cluster, for which an upper limit of $4.58\times 10^{-9}$ ph. cm$^{-2}$ s$^{-1}$ in the 0.2 to 100 GeV energy band was reported. This limit is expected to improve as the exposure is increased. In this work an updated analysis which includes additional data taken during the period 2010 February 5 to 2011 June 10 is presented as a compliment to the VERITAS results.

The analysis of this work follows the same procedure as described in detail in \citet{article:Abdo_etal:2009}. To eliminate albedo gamma rays from the Earth limb, a zenith angle cut of 105$^{\circ}$ is applied. The \emph{Fermi} Science Tools version v9r23p1 and Instrumental Response Functions (IRFs) P6\_V11\_Diffuse (a model of the spatial distribution of photon events calibrated ...) were used throughout this work. Only photons above 1 GeV are considered, where the constraints on the flux are most meaningful. The 99\% confidence level flux upper limits in the energy range 1-100 GeV, assuming both a unresolved, point-like or spatially extended emission are tabulated in Table \ref{tabel:fermi}. The gamma-ray emission is assumed to follow a power-law distribution with a photon index $\alpha=2.1$ and $2.5$.

@@

Table \ref{table:results} lists the upper limits for the selected regions of interest tabulated in Table \ref{table:roi}. These upper-limit calculations depend on the gamma-ray spectrum, which in this work is assumed to be a power law with a spectral index $\alpha=2.1, 2.3$, and $2.5$.

@@

\begin{figure*}
\begin{center}
\scalebox{0.45}{\plotone{f1a.eps}}
\scalebox{0.45}{\plotone{f1b.eps}}
\end{center}
\caption{\emph{Left}: Smoothed significance map of the Coma cluster calculated from the observed excess VHE gamma-ray events over a $4.5^{\circ}\times 4.5^{\circ}$ field of view. The excess counts were derived using a ring background model \citep{article:Aharonian_etal:2001}. White contours show the X-ray counts per second in the 0.1 to 2.4 keV energy band from the ROSAT all-sky survey \citep{article:BrielHenryBohringer:1992}. \emph{Right}: Same as above but with overlaid contours from GBT radio observations at 1.4 GHz \citep{article:BrownRudnick:2010} where strong point sources have been subtracted. Shown are also the $0.2^{\circ}$ and $0.4^{\circ}$ radii (dashed cyan) considered for the extended source analyses presented here.}
\label{fig:skymaps}
\end{figure*}

\begin{figure*}
\begin{center}
\scalebox{1.0}{\plotone{f2.eps}}
\end{center}
\caption{$\theta^{2}$ distribution from VERITAS observations of the Coma cluster of galaxies. The points with error bars represent the ON-source data sample and the filled area is the estimated background due to cosmic rays. Each bin represents an annulus around the Coma cluster core position. The data were derived from the ring background model using a $0.2^{\circ}$ sampling radius.}
\label{fig:thetasq}
\end{figure*}

@@

\begin{deluxetable}{lcc}
\tablecolumns{3}
\tablewidth{0pc}
\tablecaption{Regions of interest in the Coma cluster field of view.}
\tablehead{
\colhead{Source} &
\colhead{RA} &
\colhead{Dec}}
\startdata
Core & \RA{12}{59}{48.7} & \Dec{+27}{58}{50.0}\\
NGC 4889 & \RA{13}{00}{08.13} & \Dec{+27}{58}{37.03}\\
NGC 4874 & \RA{12}{59}{35.71} & \Dec{+27}{57}{33.37}\\
NGC 4921 & \RA{13}{01}{26.12} & \Dec{+27}{53}{09.59}\\
\enddata
\tablenotetext{a}{Note. The cluster core is considered both as a point source and a modestly extended source. Three central galaxies are also considered in point-source searches. The choice is based on evidence of an excess of non-thermal X-ray emission \citep{article:Neumann_etal:2003} at the location of these galaxies.}
\label{table:roi}
\end{deluxetable}

\begin{landscape}
\begin{deluxetable}{lcccccccccccc}
\tablecolumns{6}
\tablewidth{0pc}
\tablecaption{Flux upper limits for regions of interest in the Coma cluster of galaxies and surroundings.}
\tablehead{
\colhead{Source} & \colhead{RoI\tablenotemark{a} [deg]} & \colhead{$N_{S}$} & \colhead{S\tablenotemark{b} [$\sigma$]} &
\multicolumn{6}{c}{Flux U.L.\tablenotemark{c}} &
\multicolumn{3}{c}{Minimum gamma-ray flux\tablenotemark{d}}\\
\colhead{} & \colhead{} & \colhead{} & \colhead{} & \multicolumn{2}{c}{$\alpha=2.1$} &
\multicolumn{2}{c}{$\alpha=2.3$} &
\multicolumn{2}{c}{$\alpha=2.5$} &
\colhead{$\alpha=2.1$} &
\colhead{$\alpha=2.3$} &
\colhead{$\alpha=2.5$}}
\startdata
Core & 0 & 17 & 0.84 & 2.59 & (0.78\%) & 2.78 & (0.83\%) & 2.97 & (0.89\%) \\
& 0.2 & -41 & -1.0 & 1.96 & (0.59\%)  & 2.09 & (0.63\%) & 2.21 & (0.66\%) \\
& 0.4 & -26 & -0.30 & 4.44 & (1.3\%)  & 4.74 & (1.4\%) & 5.02 & (1.5\%)\\
NGC 4889 & 0 & 3 & 0.14 & - & - & 1.85 & (0.55\%)  & - & - \\
NGC 4874 & 0 & -14 & -0.71 & - & - &  1.51 & (0.45\%)  & - & - \\
NGC 4921 & 0 & -4 & -0.23 & - & - &  2.41 & (0.72\%)  & - & -
\enddata
\tablenotetext{a}{Intrinsic source radius (zero means point source), which is convolved with the analysis PSF.}
\tablenotetext{b}{Statistical significance calculated according to \citet{article:LiMa:1983}.}
\tablenotetext{c}{99\% confidence level upper limit in units of $10^{-8}$ ph. m$^{-2}$ s$^{-1}$ calculated according to \citet{article:Rolke_etal:2005} above an energy threshold of 220 GeV, with corresponding fluxes in percent of the steady Crab Nebula flux in parenthesis.}
\tablenotetext{d}{Minimum gamma-ray flux in units of $10^{-9}$ ph. m$^{-2}$ s$^{-1}$ from the hadronic model described in the text.}
\label{table:results}
\end{deluxetable}
\end{landscape}
